Qualitative Probabilistic Programming

### 2015-02-25 paper/paper.tex
kay

  \begin{abstract}
    Probabilistic programs with unknown functions
  \end{abstract}

@@

  Probabilistic programming languages have made it easier to specify Bayesian
  models as programs.  Once a probabilistic model is written as a probabilistic
  program, it is possible to use generic inference algorithms (such as
  Metropolis Hastings) to perform inference.
    - Church example

  In this paper, we discuss a feature of probabilistic programming languages
  that makes it easy to specify models containing unknown functions.
  Specifically, we allow programs to specify that some random functions
  are unknown.  The system will fill in these unknown functions with
  some reasonable default class of distributions based on the types and
  proceed to infer the parameters to this distribution using the
  expectation maximization algorithm.

@@

@@

  \section{Algorithm}

@@

  - EM algorithm
    - E step is Metropolis Hastings as in probabilistic programming
    - M step is parameter optimization
  - Types
    - Brief overview of exponential families
    - Gaussian
    - Categorical
    - Other exponential families could be added easily
    - Some algebraic data types
      - Tuples
      - Either
  - Conversion to factor graph
    - Different types of nodes (exponential family)
    - Different types of factors
      - Constant factors
      - Unifying values (e.g. conditioning on compound value equaling something)
      - Factors from known random functions (e.g. gaussian)
      - Factors from unknown function calls
    - Example: clustering conversion to factor graph
  - Inference in factor graph
    - Metropolis hastings
    - Proposal distribution inferred from nearby nodes

>>>>

  \begin{abstract}
    In probabilistic programs, sometimes it is difficult to specify the correct
    parameterized family of distributions.  We explore an extension to
    probabilistic programming languages that allows programmers to mark some
    distributions as unspecified.  Then, we can fill in the distribution with
    some family and infer parameters.
  \end{abstract}

@@

  Probabilistic programming languages have made it easier to specify Bayesian
  models as programs.  Once a model is written as a probabilistic
  program, it is possible to use generic inference algorithms (such as
  Metropolis Hastings) to perform inference.

  In this paper, we discuss a feature of probabilistic programming languages
  that makes it easier to specify some models.
  Specifically, we allow the programmer to specify that some random functions
  are unknown.  The system will fill in these unknown functions with
  some reasonable default family of distributions based on the types and
  proceed to infer the parameters to this distribution using the
  expectation maximization algorithm.

@@

  In an ordinary probabilistic programming language (such as Church),
  it would be possible to treat parameters as random variables.  This
  would allow ordinary inference algorithms to infer parameters.  However,
  there are advantages of having unknown functions as a feature
  in the language.
  First, it is easier to
  write programs without knowing the details of different distributions.
  Second, specialized algorithms can be used to make parameter inference
  faster.

  Inference in these models is performed using the expectation-maximization
  algorithm, with alternating steps of inferring latent variables and optimizing
  parameters.

@@

  Here is an example of a Quipp program to perform clustering:

  \begin{verbatim}
  PointType = Vector(2, Double)
  ClusterType = Categorical(3)

  get_cluster = rand_function(ClusterType)
  get_point = rand_function(ClusterType, PointType)

  def sample():
    cluster = get_cluster()
    return (cluster, get_point(cluster))
  \end{verbatim}

  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and two random functions (\texttt{get\_cluster} and \texttt{get\_point}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We do not know
  the distribution over clusters or over points given a cluster.
  We will fill in the \texttt{get\_cluster} function with
  a learned categorical distribution, and we will fill in the
  \texttt{get\_point} function with a learned unknown linear function with noise.

  \section{Class of Distributions}

    For unknown functions, the class of functions used is a kind
    of generalized linear model.  We assume that the result
    is distributed from some exponential family.  We label
    some subset of the sufficient statistics as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector of the
    features of $x$.  Then

    $$p_{\mathbf{N}}(y | x) = \exp(\psi(x)^T N \phi(y) - g(N^T \psi(X)))$$

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem:

    \begin{align*}
      \log p_{\mathbf{N}}(y | x) = \psi(x)^T N \phi(y) - g(N^T \psi(X))
    \end{align*}
    since the log probability function is concave when $g$ is convex (as it is
    for any exponential family distribution).

    Therefore, it is possible to use Newton's method to optimize the parameters.

  \section{Inference}

@@

    In the factor graph, there are 2 types of variables: real numbers
    and categorical values.
    In Quipp, there are more value types than this.  Specifically,
    in addition to real numbers and categorical types, tuples
    are also supported.
    To translate a Quipp program to a factor graph, we
    map each value to a list of variables, and create factors
    over all these variables.

    There are a few different types of factors:
    \begin{enumerate}
      \item Factors representing known distributions
      \item Factors representing unknown distribution
      \item Factors asserting that a value is equal to some constant
    \end{enumerate}

    How is inference in the factor graph performed?  We use the Metropolis Hastings
    algorithm, updating one node at a time.  To get a proposal for a given node,
    we approximate the local distribution using message passing (similar to
    in variational message passing).  Specifically, for each factor
    $f(x_1, x_2, ..., x_n)$ and each $1 \leq i \leq n$, we send a message
    to variable $x_i$ that approximates $x_i \mapsto f(x_1, x_2, ..., x_n)$.
